---
title: "RNA-seq pipeline (single-end reads)"
author: "LLi"
date: "7/12/2019"
output: html_document
---
<style type="text/css">

body{ /* Normal  */
      font-size: 16px;
  }
td {  /* Table  */
  font-size: 12px;
}
h1.title {
  font-size: 28px;
}
h1 { /* Header 1 */
  font-size: 24px;
  color: DarkBlue;
}
h2 { /* Header 2 */
    font-size: 20px;
}
h3 { /* Header 3 */
    font-size: 18px;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
library(tidyverse)
library(knitr)
library(kableExtra)
```

## Download fastq files from NCBI with SRR accession number
The RNA-seq data is from [*Ackermann AM et al.,*](https://www.ncbi.nlm.nih.gov/pubmed/26977395) "Integration of ATAC-seq and RNA-seq identifies human alpha cell and beta cell signature genes.", Mol Metab, 2016 Mar;5(3):233-244.

First download summary information regarding seq data by searching GEO accession number **GSE76268** in [NCBI SRA website](https://www.ncbi.nlm.nih.gov/sra).

R:
```{r}
# Download Summary and RunInfo csv files of seq data from NCBI SRA website manually

# Read csv files in R and display with scroller
sra_result <- read.csv("~/seq_tutorials/sra_result.csv")

kable(sra_result, caption = "sra_result.csv snapshot") %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "150px")

SraRunInfo <- read.csv("~/seq_tutorials/SraRunInfo.csv")

kable(SraRunInfo, caption = "SraRunInfo.csv snapshot") %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "150px")

# Make two column names of these two csv files the same
names(sra_result)[names(sra_result) == "Sample.Accession"] <- "Sample"
names(sra_result)[names(sra_result) == "Experiment.Accession"] <- "Experiment"

# Merge the csv files by Sample and Experiment
GSE76268_merge <- merge(sra_result, SraRunInfo, by = c("Sample", "Experiment"))

# Remove columns with NA
GSE76268_runInfo <- GSE76268_merge[colSums(!is.na(GSE76268_merge)) > 0]

# Subset with columns that provides info to concatenate fastq files
cat_runInfo <- select(GSE76268_runInfo, c("Sample", "Run", "Experiment.Title"))

kable(cat_runInfo, caption = "Information for concatenating RNA-seq fastq files") %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "150px")

```
```{r eval=FALSE}
# Save csv files
write.csv(GSE76268_runInfo, "GSE76268_runInfo.csv")
write.csv(cat_runInfo, "cat_runInfo.csv")

```

Then download SRA files to Linux server.

Shell:
```{bash eval=FALSE}
# Create a new directory for fastq files
mkdir input
cd input

# Download SRA files from NCBI (SRR3048055 to SRR3048083)
# Attention: SRA file download takes a lot of time, therefore run it in the background by screen or nohup
for (( i = 55; i <= 83; i++ ))
do
echo "Start downloading SRR30480$i"
wget https://sra-download.ncbi.nlm.nih.gov/traces/sra37/SRR/002976/SRR30480$i
done

# Check integrity of downloaded SRA file
vdb-validate SRR3048055

# fasterq-dump single-end seq data (indicate local path)
ls SRR* | while read id
do
echo "fasterq-dump on $id"
fasterq-dump -e 16 -p ./$id
done

# Remove SRA files
find . -type f  ! -name "*.*" -delete
```

Check the information in csv files and merge the fastq files that belongs to the same sample

Shell:
```{bash eval=FALSE}
# Merge files 
for file in SRR3048080* SRR3048081* SRR3048082*
do
echo "Merge $file"
cat "$file" >> ../SRS1219077.fastq && rm "$file"
done

for file in SRR3048077* SRR3048078* SRR3048079*
do
echo "Merge $file"
cat "$file" >> ../SRS1219078.fastq && rm "$file"
done

for file in SRR3048071* SRR3048072* SRR3048073*
do
echo "Merge $file"
cat "$file" >> ../SRS1219082.fastq && rm "$file"
done

for file in SRR3048067* SRR3048068* SRR3048069*
do
echo "Merge $file"
cat "$file" >> ../SRS1219084.fastq && rm "$file"
done

for file in SRR3048064* SRR3048065* SRR3048066*
do
echo "Merge $file"
cat "$file" >> ../SRS1219085.fastq && rm "$file"
done

for file in SRR3048061* SRR3048062* SRR3048063*
do
echo "Merge $file"
cat "$file" >> ../SRS1219086.fastq && rm "$file"
done

for file in *56* *57* *58*
do
echo "Merge $file"
cat "$file" >> ../SRS1219089.fastq && rm "$file"
done

# Compress fastq files (time-consuming)
parallel --gnu gzip ::: *fastq
```

## FastQC on raw sequence data

Shell:
```{bash eval=FALSE}
# Create a new directory
mkdir output_fastqc

# FastQC 
fastqc -t 32 -o ../output_fastqc *.fastq.gz

# Use MultiQC to parse summary statistics 
multiqc ../output_fastqc -o ../output_fastqc
```

## Trim adaptors

```{bash eval=FALSE}
# Warning: ONLY use GNU parallel when nobody else is using the server! (trimming uses a lot of CPU!!!)
# FastQC  will be performed after trimming
parallel trim_galore --illumina --fastqc ::: *.fastq.gz

# MultiQC
multiqc .
```

## STAR alginment
First process the human reference genome FASTA file into an index that allows the aligner to access.

```{bash eval=FALSE}
# Download human reference genome GRCh38 FASTA file
curl -O ftp://ftp.ensembl.org/pub/release-97/fasta/homo_sapiens/dna/Homo_sapiens.GRCh38.dna.toplevel.fa.gz
gunzip Homo_sapiens.GRCh38.dna.toplevel.fa.gz

# Download gene model annotation GTF file from the same release 
curl -O ftp://ftp.ensembl.org/pub/release-97/gtf/homo_sapiens/Homo_sapiens.GRCh38.97.gtf.gz
gunzip Homo_sapiens.GRCh38.97.gtf.gz

# Create STAR genome index
# All samples have sequences of a single length (100bp).
STAR --runThreadN 16 \
--limitGenomeGenerateRAM 152003700778 \
--runMode genomeGenerate \
--genomeDir /data/home/lingzili/GRCh38_genome/ \
--genomeFastaFiles /data/home/lingzili/GRCh38_genome/Homo_sapiens.GRCh38.dna.toplevel.fa \
--sjdbGTFfile /data/home/lingzili/GRCh38_genome/Homo_sapiens.GRCh38.97.gtf \
--sjdbOverhang 99

```

Then use STAR (v2.7.1a) to map reads to the hg38 genome. 
```{bash eval=FALSE}
# Align single-end reads with STAR
ls *.fq.gz |while read id
do baseFilename=`basename $id _trimmed.fq.gz`
STAR --genomeDir /data/Genomes/human/hg38/star/index_with_100bp_overhang/ \
--runThreadN 32 \
--readFilesIn $id \
--readFilesCommand zcat \
--outFileNamePrefix /data/home/lingzili/Islet_Public_Dataset/Human_bulk_RNA/GSE76268/output/STAR_hg38/${baseFilename}_ \
--outBAMsortingThreadN 16 \
--outSAMtype BAM SortedByCoordinate \
--outSAMunmapped Within
echo "${baseFilename} is aligned"
done
```

## Quality control of mapped reads
### Alignment summary
It is expected that more than 70% of the reads should uniquely map to the reference genome. Contamination, low quality sequencing data, or poor library contruction may result in <60% uniquely aligned reads.

### Insert size (Picard)

### rRNA percentage
Ribosomal RNA (rRNA) often exceeding 90% of the total RNA. It is expected that prior library treatment of rRNA removal should result in less than 5% of the reads mapped to rRNA sequences.


### Read distribution
It is expected that distribition profiles should be similar across samples. Less than 20% intronic mapped reads are expected in mRNA-seq libraries, whereas more than 40% in total RNA-seq libraries.


### Gene body coverage
Uniform distributed reads across genebodies is a prerequisite for accurate quantification of gene expression. The line plots should look like this: "∩". mRNA-seq library preparation typically uses oligo-dT beads to capture mature transcripts and can be prone to 3' bias in genebody coverage if degraded RNA (RIN < 7) is used as input, which may potentially introduce quantification bias.



### Junction saturation


## Differential gene expression using DESeq2
We considered genes with an FDR<.1 as significantly differentially expressed.

### Correlation Heatmap
**Spearman correlation heatmap between biological replicates**

**Heatmap with pre-defined genes**
With 50 most variable genes

### Volcano plot

### GOTerm analysis with GOstats

### KEGG pathway with GAGE and clusterProfiler

### Gene discovery rate (detected genes vs. reads)
See <Gene Discovery Rate in RNA-seq data>(http://www.nxn.se/valent/gene-discovery-rate-in-rna-seq-data)

## Unsupervised clustering
### PCA
Principle component 1 (PC1) accounts for the most amount of variation across samples, and PC2 the second most.

## Optional: Batch correction