---
title: 'Coding club meet-up: Using ggplot2 for visualization'
author: "Lingzi Li"
date: "December 3 2019"
output: html_document
---
<style type="text/css">

body{ /* Normal  */
      font-size: 16px;
  }
td {  /* Table  */
  font-size: 12px;
}
h1.title {
  font-size: 28px;
}
h1 { /* Header 1 */
  font-size: 24px;
  color: DarkBlue;
}
h2 { /* Header 2 */
    font-size: 20px;
}
h3 { /* Header 3 */
    font-size: 18px;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, python.reticulate = FALSE)
```

# The Anatomy of a plot

## Load data

```{r Load packages, echo=TRUE}
# Load packages
library(tidyverse)
library(readxl)
```

```{r Load data, include=FALSE}
# Load data
Mass_data <- read_excel("~/biology_graph/data/Gcg_DEXA.xlsx")

# Gather columns into key-value pairs
Mass_data_long <- Mass_data %>%
  select(-c(Total_gram)) %>%
  gather(Composition, Gram, 6:7)

# Remove "_gram" from data labels
Mass_data_long$Composition <- gsub("_gram", "", Mass_data_long$Composition)

# Rank the diet
Mass_data_long$Diet <- factor(Mass_data_long$Diet, levels = c("Chow", "HFD"))

# Subset fat mass of male
fatMass_Male <- Mass_data_long %>%
  filter(Sex == "Male", Composition == "Fat")
```

```{r Show table, echo=TRUE}
head(fatMass_Male)
```

## 1) Plot background

```{r Plot background, echo=TRUE, fig.height=3, fig.width=6}
# Plot fat mass of male
p1 <- fatMass_Male %>%
  ggplot(aes(x = Diet, y = Gram, fill = Diet))

p1
```

## 2) First layer

```{r First layer, echo=TRUE, fig.height=3, fig.width=6}
p2 <- p1 + geom_boxplot()

p2
```

## 3) Another layer

```{r Another layer, echo=TRUE, fig.height=3, fig.width=6}
p3 <- p2 + geom_jitter(width = .1, size = 3)

p3
```

## 4) Lay out panels in a grid

```{r Lay out panels, echo=TRUE, fig.height=3, fig.width=6}
p4 <- p3 + facet_grid(cols = vars(Diet_duration))

p4
```

# 5) Complete with a customized look

```{r Plot theme, echo=TRUE, fig.height=4, fig.width=6}
My_theme <- theme(
  axis.line = element_line(colour = "black"),
  axis.text.x = element_text(color = "black", size = 14, face = "bold"),
  axis.text.y = element_text(color = "black", size = 14, face = "bold"),
  axis.title.x = element_text(color = "black", size = 14, face = "bold"),
  axis.title.y = element_text(color = "black", size = 14, face = "bold"),
  strip.text.x = element_text(color = "black", size = 14, face = "bold"), # Horizontal facet labels
  strip.background = element_rect(fill = "white"), # Background of facet labels
  panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  panel.background = element_blank(),
  panel.border = element_rect(colour = "black", fill = NA, size = 2),
  plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"),
  plot.title = element_text(color = "black", size = 16, face = "bold")
)

p5 <- p4 + My_theme

p5
```

# Geometric objects (geoms)

## Dot plot with fitted line

### Dataset: Standard curve of insulin ELISA, performed on different dates

```{r ELISA data, echo=FALSE}
Ins_Cal <- read.csv("~/assay_calculation/data/Cal_Ins.csv")

Ins_Cal_v2 <- Ins_Cal %>%
  gather(Date, Intensity, 3:10)

# Remove "X" from data labels
Ins_Cal_v2$Date <- gsub("X", "", Ins_Cal_v2$Date)

head(Ins_Cal_v2, n = 10L)
```

### 1) Plot background of standard curve

```{r Dot plot p1, echo=TRUE, fig.height=3, fig.width=6}
Ins_p1 <- Ins_Cal_v2 %>%
  ggplot(aes(x = Insulin, y = Intensity))

Ins_p1
```

### 2) Add individual data points

```{r Dot plot p2, echo=TRUE, fig.height=3, fig.width=6}
Ins_p2 <- Ins_p1 + geom_point(size = 3)

Ins_p2
```

### 3.1) Add smoothed conditional means with standard error

```{r Dot plot p3, echo=TRUE, fig.height=3, fig.width=6}
Ins_p3 <- Ins_p2 + geom_smooth(span = 0.8, method = "loess", formula = y ~ x)

Ins_p3
```

### 3.2) Add a line of best fit

```{r Dot plot p4, echo=TRUE, fig.height=3, fig.width=6}
Ins_p4 <- Ins_p2 + geom_smooth(method = "lm", se = FALSE)

Ins_p4
```

```{r Dot plot lm, echo=TRUE, fig.height=3, fig.width=6}
# Linear regression
# Intensity = Intercept + (ß * Insulin)
linearMod <- lm(Intensity ~ Insulin, data = Ins_Cal_v2)
summary(linearMod)
```

## Bar chart

### Dataset (same as above): Standard curve of insulin ELISA, performed on different dates

### 1) Plot background of bar chart

```{r bar chart p1, echo=TRUE, fig.height=3, fig.width=6}
Ins_Bar_p1 <- Ins_Cal_v2 %>%
  ggplot(aes(x = Calibrator, y = Intensity, fill = Date))

Ins_Bar_p1
```

### 2) Add bars in groups

```{r bar chart p2, echo=TRUE, fig.height=3, fig.width=6}
Ins_Bar_p2 <- Ins_Bar_p1 + geom_bar(stat = "identity", position = position_dodge())

Ins_Bar_p2
```

### 3) Add bars with means with standard error

```{r Ins mean and se, echo=TRUE}
# Calculate mean and se
Ins_Stat <- Ins_Cal_v2 %>%
  group_by(Insulin, Calibrator) %>%
  summarise(
    average = mean(Intensity, na.rm = TRUE),
    se = sd(Intensity, na.rm = TRUE) / sqrt(length(Intensity))
  )

Ins_Stat
```

```{r bar chart p3, echo=TRUE, fig.height=3, fig.width=6}
# First plot the mean
Ins_Bar_p3 <- Ins_Stat %>%
  ggplot(aes(x = Calibrator, y = average, fill = Calibrator)) +
  geom_bar(stat = "identity")

Ins_Bar_p3
```

```{r bar chart p4, echo=TRUE, fig.height=3, fig.width=6}
# Then add error bars
Ins_Bar_p4 <- Ins_Bar_p3 +
  geom_errorbar(aes(ymin = average - se, ymax = average + se), width = .2) +
  labs(title = "Plot of insulin calibrator", x = NULL, y = "Intensity A450")

Ins_Bar_p4
```

## Violin plot

### Dataset (same as above): Standard curve of insulin ELISA, performed on different dates

### Basic violin plot

```{r violin p1, echo=TRUE, fig.height=3, fig.width=6}
Ins_Violin_p1 <- Ins_Cal_v2 %>%
  ggplot(aes(x = Calibrator, y = Intensity, color = Calibrator)) +
  geom_violin()

Ins_Violin_p1
```

```{r violin p1 rotate, echo=TRUE, fig.height=3, fig.width=6}
# Rotate the violin plot
Ins_Violin_p1 + coord_flip()
```

### Add mean with standard deviation

```{r r violin p2, echo=TRUE, fig.height=4, fig.width=6}
Ins_Violin_p1 + stat_summary(fun.data = mean_sdl, geom = "pointrange", color = "black")
```

## Venn diagram

### Dataset for venn diagram
```{r df venn, echo=FALSE}
# Set the seed of random number generator
set.seed((100))

# Random generation for the binomial distribution with parameters size (0 or 1) and prob.
df_venn <- data.frame(
  A = rbinom(100, 1, 0.8),
  B = rbinom(100, 1, 0.7),
  C = rbinom(100, 1, 0.6)
) %>% mutate_all(., as.logical)

# Add "Gene_" to row names
rownames(df_venn) <- paste0("Gene_", rownames(df_venn))

head(df_venn)
```

Use vennCounts from the package limma to compute classification counts

```{r vennCounts, echo=TRUE}
# Load packages
library(ggforce)
library(limma)

# Compute classification counts
counts_venn <- vennCounts(df_venn)
class(counts_venn) <- "matrix"

# Add x and y coordinates for the count annotations
counts_venn <- as.data.frame(counts_venn)[-1, ] %>%
  mutate(
    x = c(0, 1.2, 0.8, -1.2, -0.8, 0, 0),
    y = c(1.2, -0.6, 0.5, -0.6, 0.5, -1, 0)
  )

counts_venn
```

### Define basic structure for the circles

```{r Define coordinates, echo=TRUE, fig.height=4, fig.width=6}
# Define x and y coordinates for the circles
venn_structure <- data.frame(
  x = c(0, 0.866, -0.866),
  y = c(1, -0.5, -0.5),
  Sample = c("A", "B", "C")
)

venn_p1 <- venn_structure %>% ggplot(aes(x0 = x, y0 = y, r = 1.5, fill = Sample)) +
  geom_circle(alpha = .5, size = 1, colour = "grey") +
  coord_fixed()

venn_p1
```

### Venn diagram with annotation of the counts

```{r Venn p2, echo=TRUE, fig.height=4, fig.width=6}
venn_p2 <- venn_p1 +
  annotate("text", x = counts_venn$x, y = counts_venn$y, label = counts_venn$Counts, size = 5)

venn_p2
```

Finally to remove the grey background

```{r venn blank background, echo=TRUE, fig.height=4, fig.width=6}
venn_p2 + theme_void()
```

## PCA plot

### Dataset from RNA-seq tpm counts

```{r Calculate pca, echo=TRUE}
# Load kallisto counts
kallisto_count <- read.csv("~/mRNA_IP/count/kallisto_count_19112019.csv")

kallisto_df <- kallisto_count[c(1, 5, 6, 7)] %>%
  spread(target_id, tpm, fill = NA, convert = FALSE, drop = TRUE, sep = NULL)

kallisto_df[1:9, 1:6]

# Calculate principal component analysis based on tpm
pca <- prcomp(kallisto_df[, -c(1, 2)])

pca$x
```

### PCA plot

Option 1: By using ggplot2

```{r pca df, echo=TRUE}
# Create data frame for PC
df_pca <- as.data.frame(pca$x)

df_pca$condition <- kallisto_df$condition

head(df_pca)
```

```{r pca ggplot2, echo=TRUE, fig.height=3, fig.width=5}
# Plot PCA with ggplot2
pca_p1 <- ggplot(df_pca, aes(x = PC1, y = PC2, color = condition))

pca_p1 + geom_point(size = 3) + theme_bw()
```

Option 2: By using package ggfortify

```{r ggfortify, echo=TRUE, fig.height=3, fig.width=5}
library(ggfortify)
pca_v2_p1 <- autoplot(pca, data = kallisto_df, colour = "condition", size = 3) +
  theme_bw()

pca_v2_p1
```

## Heatmap

### Dataset from RNA-seq tpm counts

```{r Prepare data, echo=TRUE}
df_tpm <- kallisto_count %>%
  filter(tpm > 0) %>%
  select(c(1, 5, 6))

head(df_tpm)
```

### Subset a list of interested genes

```{r Subset genelist, echo=TRUE}
# Ins1: ENSMUST00000039652.5
# Mafa: ENSMUST00000062002.5
# Pdx1: ENSMUST00000085591.6
# Gcg: ENSMUST00000102733.9
# Sst: ENSMUST00000004480.4
# Amy1: ENSMUST00000106540.7
# Pnlip: ENSMUST00000057270.8

geneList <- c("ENSMUST00000039652.5", "ENSMUST00000062002.5", "ENSMUST00000085591.6", "ENSMUST00000102733.9", "ENSMUST00000004480.4", "ENSMUST00000106540.7", "ENSMUST00000057270.8")

# Subset interested genes
df_subset <- subset(df_tpm, df_tpm$target_id %in% geneList)

# Calculate z scores
df_subset$z <- runif(df_subset$tpm, min = -10, max = 10)

head(df_subset)
```

```{r heatmap, echo=TRUE, fig.height=5, fig.width=6}
# Set the theme for heatmap
theme_heatmap <- theme(
  axis.line = element_line(colour = "black"),
  axis.text.x = element_text(color = "black", size = 12, face = "bold", angle = 45, hjust = 1),
  axis.text.y = element_text(color = "black", size = 16, face = "bold"),
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  legend.title = element_blank(),
  legend.text = element_text(color = "black", size = 16, face = "bold"),
  legend.key = element_rect(fill = "white"), # Remove grey background of the legend
  panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  panel.background = element_blank(),
  panel.border = element_rect(colour = "black", fill = NA, size = 2),
  plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm")
)

# heatmap by geom_tile
df_subset %>% ggplot(aes(x = sample, y = target_id)) +
  geom_tile(aes(fill = z)) +
  scale_fill_gradient(low = "white", high = "red") +
  theme_heatmap +
  scale_y_discrete(labels = c("ENSMUST00000039652.5" = "Ins1", "ENSMUST00000062002.5" = "Mafa", "ENSMUST00000085591.6" = "Pdx1", "ENSMUST00000102733.9" = "Gcg", "ENSMUST00000004480.4" = "Sst", "ENSMUST00000106540.7" = "Amy1", "ENSMUST00000057270.8" = "Pnlip"))
```

# References

ggplot2: Elegant Graphics for Data Analysis by Hadley Wickham <https://ggplot2-book.org/>

ggplot2 - Essentials <http://www.sthda.com/english/wiki/ggplot2-essentials>
