---
title: "How to upgrade R on Windows"
author: "LLi"
date: "7/9/2019"
output: html_document
---

<style type="text/css">

h1.title {
  font-size: 30px;
  color: Green;
}

h1 { /* Header 1 */
  font-size: 24px;
  color: DarkBlue;
}

body{ /* Normal  */
      font-size: 16px;
  }

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Step 1: Check the current R version on your PC

You can get the information of the current version of R on your PC by either way:

- Open RStudio and you will find session information on the top of the Console

- Type in the Console:

```{r}
R.version.string
```

# Step 2: Check the latest R release

Go to official R website <https://www.r-project.org/>

# Step 3: Use installr package to upgrade R

First, install installr package (see Documentation <https://www.rdocumentation.org/packages/installr>):

```{r eval=FALSE}
install.packages("installr")
```

Then, load the package:

```{r eval=FALSE}
library("installr")
```

Use the function updateR (see Documentation <https://www.rdocumentation.org/packages/installr/versions/0.21.3/topics/updateR>) with specified arguments:

```{r eval=FALSE}
updateR(
  browse_news = FALSE, # If you are not interested in seeing the news of the latest release
  copy_packages = TRUE, # Copy your packages from the old version to the new version of R
  keep_old_packages = TRUE, # Keep the packages in the library of the old R installation
  update_packages = TRUE, # Update your packages in the new version of R
  keep_install_file = TRUE # Keep installer file after download
)
```

If you need to log in to gain administrator rights for installation, it is possible that updateR will pause and give an error code after downloading the installer file. 

Don't panic! Here is how you can get out of this problem:

- Read the error message and find out the path where the installer file is installed on your PC

- Use the indicated path to locate the the installer file that ends with .exe on your PC

- Right click the installer file and select **Run as administrator**

- Log in to gain administrator rights and follow the prompts to complete installation
