---
title: "RNA-seq pipeline (single-end reads)"
author: "LLi"
date: "7/12/2019"
output: html_document
---
<style type="text/css">

body{ /* Normal  */
      font-size: 16px;
  }
td {  /* Table  */
  font-size: 12px;
}
h1.title {
  font-size: 28px;
}
h1 { /* Header 1 */
  font-size: 24px;
  color: DarkBlue;
}
h2 { /* Header 2 */
    font-size: 20px;
}
h3 { /* Header 3 */
    font-size: 18px;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
library(tidyverse)
library(knitr)
library(kableExtra)
```

## Download fastq files from NCBI with SRR accession number
The RNA-seq data is from [*Ackermann AM et al.,*](https://www.ncbi.nlm.nih.gov/pubmed/26977395) "Integration of ATAC-seq and RNA-seq identifies human alpha cell and beta cell signature genes.", Mol Metab, 2016 Mar;5(3):233-244.

First download summary information regarding seq data by searching GEO accession number **GSE76268** in [NCBI SRA website](https://www.ncbi.nlm.nih.gov/sra).

R:
```{r}
# Download Summary and RunInfo csv files of seq data from NCBI SRA website manually

# Read csv files in R and display with scroller
sra_result <- read.csv("~/seq_tutorials/data/sra_result.csv")

kable(sra_result, caption = "sra_result.csv snapshot") %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "150px")

SraRunInfo <- read.csv("~/seq_tutorials/data/SraRunInfo.csv")

kable(SraRunInfo, caption = "SraRunInfo.csv snapshot") %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "150px")

# Make two column names of these two csv files the same
names(sra_result)[names(sra_result) == "Sample.Accession"] <- "Sample"
names(sra_result)[names(sra_result) == "Experiment.Accession"] <- "Experiment"

# Merge the csv files by Sample and Experiment
GSE76268_merge <- merge(sra_result, SraRunInfo, by = c("Sample", "Experiment"))

# Remove columns with NA
GSE76268_runInfo <- GSE76268_merge[colSums(!is.na(GSE76268_merge)) > 0]

# Subset with columns that provides info to concatenate fastq files
cat_runInfo <- select(GSE76268_runInfo, c("Sample", "Run", "Experiment.Title"))

kable(cat_runInfo, caption = "Information for concatenating RNA-seq fastq files") %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "150px")

```
```{r eval=FALSE}
# Save csv files
write.csv(GSE76268_runInfo, "./data/GSE76268_runInfo.csv")
write.csv(cat_runInfo, "./data/cat_runInfo.csv")

```

Then download SRA files to Linux server.

Shell:
```{bash eval=FALSE}
# Create a new directory for fastq files
mkdir input
cd input

# Download SRA files from NCBI (SRR3048055 to SRR3048083)
# Attention: SRA file download takes a lot of time, therefore run it in the background by screen or nohup
for (( i = 55; i <= 83; i++ ))
do
echo "Start downloading SRR30480$i"
wget https://sra-download.ncbi.nlm.nih.gov/traces/sra37/SRR/002976/SRR30480$i
done

# Check integrity of downloaded SRA file
vdb-validate SRR3048055

# fasterq-dump single-end seq data (indicate local path)
ls SRR* | while read id
do
echo "fasterq-dump on $id"
fasterq-dump -e 16 -p ./$id
done

# Remove SRA files
find . -type f  ! -name "*.*" -delete
```

Check the information in csv files and merge the fastq files that belongs to the same sample

Shell:
```{bash eval=FALSE}
# Merge files 
for file in SRR3048080* SRR3048081* SRR3048082*
do
echo "Merge $file"
cat "$file" >> ../SRS1219077.fastq && rm "$file"
done

for file in SRR3048077* SRR3048078* SRR3048079*
do
echo "Merge $file"
cat "$file" >> ../SRS1219078.fastq && rm "$file"
done

for file in SRR3048071* SRR3048072* SRR3048073*
do
echo "Merge $file"
cat "$file" >> ../SRS1219082.fastq && rm "$file"
done

for file in SRR3048067* SRR3048068* SRR3048069*
do
echo "Merge $file"
cat "$file" >> ../SRS1219084.fastq && rm "$file"
done

for file in SRR3048064* SRR3048065* SRR3048066*
do
echo "Merge $file"
cat "$file" >> ../SRS1219085.fastq && rm "$file"
done

for file in SRR3048061* SRR3048062* SRR3048063*
do
echo "Merge $file"
cat "$file" >> ../SRS1219086.fastq && rm "$file"
done

for file in *56* *57* *58*
do
echo "Merge $file"
cat "$file" >> ../SRS1219089.fastq && rm "$file"
done

# Compress fastq files (time-consuming)
parallel --gnu gzip ::: *fastq
```

## FastQC on raw sequence data

Shell:
```{bash eval=FALSE}
# Create a new directory
mkdir output_fastqc

# FastQC 
fastqc -t 32 -o ../output_fastqc *.fastq.gz

# Use MultiQC to parse summary statistics 
multiqc ../output_fastqc -o ../output_fastqc
```

## Trim adaptors

```{bash eval=FALSE}
# Warning: ONLY use GNU parallel when nobody else is using the server! (trimming uses a lot of CPU!!!)
# FastQC  will be performed after trimming
parallel trim_galore --illumina --fastqc ::: *.fastq.gz

# MultiQC
multiqc .
```

## STAR alginment
First process the human reference genome FASTA file into an index that allows the aligner to access.

```{bash eval=FALSE}
# Download human reference genome GRCh38 FASTA file
curl -O ftp://ftp.ensembl.org/pub/release-97/fasta/homo_sapiens/dna/Homo_sapiens.GRCh38.dna.toplevel.fa.gz
gunzip Homo_sapiens.GRCh38.dna.toplevel.fa.gz

# Download gene model annotation GTF file from the same release 
curl -O ftp://ftp.ensembl.org/pub/release-97/gtf/homo_sapiens/Homo_sapiens.GRCh38.97.gtf.gz
gunzip Homo_sapiens.GRCh38.97.gtf.gz

# Create STAR genome index (Lasted for 6 hours!!)
# All samples have sequences of a single length (100bp).
STAR --runThreadN 16 \
--limitGenomeGenerateRAM 152003700778 \
--runMode genomeGenerate \
--genomeDir /data/home/lingzili/GRCh38_genome/ \
--genomeFastaFiles /data/home/lingzili/GRCh38_genome/Homo_sapiens.GRCh38.dna.toplevel.fa \
--sjdbGTFfile /data/home/lingzili/GRCh38_genome/Homo_sapiens.GRCh38.97.gtf \
--sjdbOverhang 99
```

Then use STAR (v2.7.1a) to map reads to the hg38 genome. 
```{bash eval=FALSE}
# Align single-end reads with STAR
ls *.fq.gz |while read id
do baseFilename=`basename $id _trimmed.fq.gz`
STAR --genomeDir /data/home/lingzili/GRCh38_genome/STAR \
--runThreadN 32 \
--readFilesIn $id \
--readFilesCommand zcat \
--outFileNamePrefix /data/home/lingzili/Islet_Public_Dataset/Human_bulk_RNA/GSE76268/analysis/STAR/${baseFilename}_ \
--outBAMsortingThreadN 32 \
--outSAMtype BAM SortedByCoordinate \
--outSAMunmapped Within # This should have been set as None (default)
echo "${baseFilename} is aligned"
done

# Index BAM files
# Only use parallel when nobody else uses the server!!
parallel "samtools index {}" ::: *bam
```

## Visulization by IGV

Download .bam and .bai file to local desktop. IGV requires the .bai file to be in the same location as the .bam file that is loaded into IGV, but there is no direct use for that file.

## Quality control of mapped reads
Quality metrics are accessed by RNASeQC, Picard and MultiQC.

```{bash eval=FALSE}
# The process needs to be optimized for parallel
bash mapQC.sh

# MultiQC all mapping stats
multiqc .
```

```{bash eval=FALSE}
# RNA-seq reads coverage over gene body over all BAM files
# Only input sorted and indexed BAM files
geneBody_coverage.py -r $housekeepBED -i $sortedIndexedBAM -o geneBody_GSE76268

```

### **Insert size (Picard)**
There is no way to calculate average fragment size based on single-end sequencing data.

### **Alignment summary**
It is expected that more than 70% of the reads should uniquely map to the reference genome. Contamination, low quality sequencing data, or poor library contruction may result in <60% uniquely aligned reads.

### **rRNA percentage**
Ribosomal RNA (rRNA) often exceeding 90% of the total RNA. It is expected that prior library treatment of rRNA removal should result in less than 5% of the reads mapped to rRNA sequences. Excess rRNA (> 2%) will normally have to be filtered out so that differences in rRNA mapped reads across samples do not affect alignment rates and skew subsequent normalization of the data.

### **Read distribution**
It is expected that distribition profiles should be similar across samples. Less than 20% intronic mapped reads are expected in mRNA-seq libraries, whereas more than 40% in total RNA-seq libraries. A high intronic mapping suggests possible genomic DNA contamination and/or pre-mRNA. Around 55% of reads are normally observed mapping to exons.

### **Gene body coverage**
Uniform distributed reads across genebodies is a prerequisite for accurate quantification of gene expression. The line plots should look like this: "∩". mRNA-seq library preparation typically uses oligo-dT beads to capture mature transcripts and can be prone to 3' bias in genebody coverage if degraded RNA (RIN < 7) is used as input, which may potentially introduce quantification bias.

### **Transcript integrity number (TIN)**
TIN is to measure RNA integrity of individual transcripts/genes. Samples with lower median TIN score of all the transcripts usually had more skewed gene body coverage.

## Generate a count matrix using featureCounts
Only uniquely mapped reads are counted at the gene level. Total read count of a gene (meta-feature) is the sum of reads mapped to each of the exons (feature).

```{bash eval=FALSE}
# Count features in all BAM files together (single-end reads)
# Required option for specifying path to, and name of the text output (count matrix)
# 2> redirects all the information from the terminal/screen into a file
GTF="/data/home/lingzili/GRCh38_genome/WholeGenomeFasta/Homo_sapiens.GRCh38.97.gtf"
outputFile="/data/home/lingzili/Islet_Public_Dataset/Human_bulk_RNA/GSE76268/analysis/featurecount/GSE76268_featureCounts.txt" 
terminalLog="/data/home/lingzili/Islet_Public_Dataset/Human_bulk_RNA/GSE76268/analysis/featurecount/GSE76268_featureCounts.output" 
featureCounts -T 16 --verbose -a $GTF -o $outputFile *.bam 2>$terminalLog

# MultiQC count stats
cd /data/home/lingzili/Islet_Public_Dataset/Human_bulk_RNA/GSE76268/analysis/featurecount
multiqc .
```

The output txt file contains columns: Geneid, Chr, Start, End, Strand, Length and names of BAM files. We need to extract Geneid and their counts in each BAM file, which are needed for downstream processing.

```{bash eval=FALSE}
# Inspect txt file
head -n 3 GSE76268_featureCounts.txt

# Extract Geneid and their counts in each BAM file 
cut -f1,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21 GSE76268_featureCounts.txt >GSE76268_rMatrix.txt
```
